Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-11-27 inst/manuscript/manuscript.Rmd
some more updates

There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site. These data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry. Moose are most prevalent in the southwestern and eastern parts of the Togiak region (Figure$~$\ref{fig-moose-data}).

@@

SPGLMs in \pkg{spmodel} for point-referenced data are fit using the \code{spglm()} function. The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the response variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance. The \code{formula}, \code{family}, and \code{data} arguments are the three required arguments to \code{glm()} for nonspatial GLMs. So, the transition from \code{glm()} to \code{spglm()}
simply requires one additional argument: \code{spcov_type}. When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent the projected x- and y-coordinates, respectively.

@@

```{r}
bin <- spglm(
  formula = presence ~ elev + strat,
  family = binomial,
  data = moose,
  spcov_type = "none"
)
```

While the \code{spglm()} approach evaluates the HGLMM likelihood with $\sigma^2_{de} = 0$ and $\sigma^2_{ie} \approx 0$ instead of just the GLM likelihood, the parameter estimates and their standard errors are the same:

```{r}
bin_glm <- glm(
  formula = presence ~ elev + strat,
  family = binomial,
  data = moose,
)
round(coef(bin), digits = 4)
round(coef(bin_glm), digits = 4)
round(sqrt(diag(vcov(bin))), digits = 4)
round(sqrt(diag(vcov(bin_glm))), digits = 4)
```

However, using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same likelihood:
```{r}
glance(spbin)
glance(bin)
```

The likelihood-based statistics AIC, AICc, BIC, and deviance are much lower for the SPGLM, indicating a better fit relative to the GLM. We may also perform a likelihood ratio test (LRT) between the two models, as the GLM is a special case of the SPGLM (i.e., is nested within the SPGLM):
```{r}
anova(spbin, bin)
```

The LRT provides strong evidence that the SPGLM is preferred to the GLM ($p~$value < 0.001).

An alternative approach to model comparison is to use a cross-validation procedure [@james2013introduction]. The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ in each iteration. Performing leave-one-out cross validation tends to be more computationally efficient than fitting the model, as leave-one-out cross validation requires only one set of products involving the inverse covariance matrix (a primary computational burden), while fitting traditional models requires these products for each optimization iteration. After performing leave-one-out cross validation, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient. Three separate metrics (likelihood-based statistics, likelihood-ratio test, and leave-one-out cross validation) prefer the SPGLM to the GLM.

We can compare two SPGLMs with different spatial covariance functions using likelihood-based statistics and leave-one-out cross validation, but we can't use the LRT because generally, the spatial covariance functions are not nested:
```{r}
spbin2 <- update(spbin, spcov_type = "gaussian")
glances(spbin, spbin2)
loocv(spbin)
loocv(spbin2)
```

The \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower (better) deviance but slightly higher (worse) AIC, AICc, and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}). Both spatial covariance functions have similar leave-one-out cross validation metrics, though the \code{"gaussian"} spatial covariance RMSPE is slightly lower (better). For practical purposes, these models fit similarly.

>>>>

There are five columns in `moose`: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels (\code{"L"} and \code{"M"}) categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site. These data have point-referenced support because each observation occurs at point coordinates represented by a \code{POINT} geometry. Moose are most prevalent in the southwestern and eastern parts of the Togiak region (Figure$~$\ref{fig-moose-data}).

@@

SPGLMs in \pkg{spmodel} for point-referenced support are fit using the \code{spglm()} function. The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the response variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance. The \code{formula}, \code{family}, and \code{data} arguments are familiar because they are the three required arguments to \code{glm()} for nonspatial GLMs. So the transition from \code{glm()} to \code{spglm()}
simply requires one additional argument: \code{spcov_type}. When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent the projected x- and y-coordinates, respectively.

@@

```{r}
bin <- spglm(
  formula = presence ~ elev + strat,
  family = binomial,
  data = moose,
  spcov_type = "none"
)

bin_glm <- glm(
  formula = presence ~ elev + strat,
  family = binomial,
  data = moose,
)
round(coef(bin), digits = 4)
round(coef(bin_glm), digits = 4)
round(sqrt(diag(vcov(bin))), digits = 4)
round(sqrt(diag(vcov(bin_glm))), digits = 4)
```

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation~\ref{eq-marginal}):
```{r}
glance(spbin)
glance(bin)
```

The likelihood-based statistics AIC, AICc, BIC, and deviance are much lower for the SPGLM, indicating a better fit relative to the GLM.

Instead of relying on likelihood-based statistics, models can be compared using a cross-validation procedure [@james2013introduction]. The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ in each iteration. Performing leave-one-out cross validation tends to be more computationally efficient than fitting the model, as leave-one-out cross validation requires only one set of products involving the inverse covariance matrix (a primary computational burden), while fitting traditional models requires these products for each optimization iteration. After performing leave-one-out cross validation, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient. Both the liklelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.

We may also want to compare the fit of two SPGLMs with different spatial covariance structures. For example, we may want to compare the fit of an SPGLM with the `"exponential"` spatial covariance to the fit of a SPGLM with the `"gaussian"` spatial covariance:
```{r}
spbin2 <- update(spbin, spcov_type = "gaussian")
glances(spbin, spbin2)
loocv(spbin)
loocv(spbin2)
```

The SPGLM with the \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower (better) deviance but slightly higher (worse) AIC, AICc, and BIC than the SPGLM with the \code{"gaussian"} spatial covariance (\code{spbin2}). Both SPGLMs have similar leave-one-out cross validation metrics, though the SPGLM with the \code{"gaussian"} spatial covariance has slightly lower (better) RMSPE. For practical purposes, these models fit similarly.

### 2025-11-27 inst/manuscript/manuscript.tex
some more updates

There are five columns: \code{elev}, the numeric site elevation
(meters); \code{strat} a stratification variable for sampling with two
levels, \code{"L"} and \code{"M"}, which are categorized by landscape
metrics at each site; \code{count}, the number of moose at each site;
\code{presence}, a factor that indicates whether at least one moose was
observed at each site (\code{0} implies no moose; \code{1} implies at
least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG:
3338) projected coordinate of each site. These data are point-referenced
because each observation occurs at point coordinates and are represented
by a \code{POINT} geometry. Moose are most prevalent in the southwestern
and eastern parts of the Togiak region
(Figure\(~\)\ref{fig-moose-data}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bin <- spglm(
+   formula = presence ~ elev + strat,
+   family = binomial,
+   data = moose,
+   spcov_type = "none"
+ )
\end{CodeInput}
\end{CodeChunk}

While the \code{spglm()} approach evaluates the HGLMM likelihood with
\(\sigma^2_{de} = 0\) and \(\sigma^2_{ie} \approx 0\) instead of just
the GLM likelihood, the parameter estimates and their standard errors
are the same:

\begin{CodeChunk}
\begin{CodeInput}
R> bin_glm <- glm(
+   formula = presence ~ elev + strat,
+   family = binomial,
+   data = moose,
+ )
R> round(coef(bin), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
    -0.4247     -0.0003      0.8070 
\end{CodeOutput}
\begin{CodeInput}
R> round(coef(bin_glm), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
    -0.4247     -0.0003      0.8070 
\end{CodeOutput}
\begin{CodeInput}
R> round(sqrt(diag(vcov(bin))), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
     0.4208      0.0019      0.2906 
\end{CodeOutput}
\begin{CodeInput}
R> round(sqrt(diag(vcov(bin_glm))), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
     0.4208      0.0019      0.2906 
\end{CodeOutput}
\end{CodeChunk}

However, using \code{spglm()} instead of \code{glm()} ensures that
\pkg{spmodel} helper functions are available and that each of the
\code{spglm()} models uses the same likelihood:

@@

The likelihood-based statistics AIC, AICc, BIC, and deviance are much
lower for the SPGLM, indicating a better fit relative to the GLM. We may
also perform a likelihood ratio test (LRT) between the two models, as
the GLM is a special case of the SPGLM (i.e., is nested within the
SPGLM):

\begin{CodeChunk}
\begin{CodeInput}
R> anova(spbin, bin)
\end{CodeInput}
\begin{CodeOutput}
Likelihood Ratio Test

Response: presence
             Df   Chi2 Pr(>Chi2)    
spbin vs bin  3 31.546 6.525e-07 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

The LRT provides strong evidence that the SPGLM is preferred to the GLM
(\(p~\)value \textless{} 0.001).

An alternative approach to model comparison is to use a cross-validation
procedure \citep{james2013introduction}. The \code{loocv()} function
performs leave-one-out cross validation, comparing the predicted mean
(on the response scale) to the observed response variable for each
hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\) in
each iteration. Performing leave-one-out cross validation tends to be
more computationally efficient than fitting the model, as leave-one-out
cross validation requires only one set of products involving the inverse
covariance matrix (a primary computational burden), while fitting
traditional models requires these products for each optimization
iteration. After performing leave-one-out cross validation, statistics
like bias, mean-squared-prediction error (MSPE), and the square root of
MSPE (RMSPE) can be used to evaluate models:

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and
RMSPE than the GLM, indicating the SPGLM predictions are far more
efficient. Three separate metrics (likelihood-based statistics,
likelihood-ratio test, and leave-one-out cross validation) prefer the
SPGLM to the GLM.

We can compare two SPGLMs with different spatial covariance functions
using likelihood-based statistics and leave-one-out cross validation,
but we can't use the LRT because generally, the spatial covariance
functions are not nested:

@@

The \code{"exponential"} spatial covariance (\code{spbin}) has a
slightly lower (better) deviance but slightly higher (worse) AIC, AICc,
and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}).
Both spatial covariance functions have similar leave-one-out cross
validation metrics, though the \code{"gaussian"} spatial covariance
RMSPE is slightly lower (better). For practical purposes, these models
fit similarly.

>>>>

There are five columns in \texttt{moose}: \code{elev}, the numeric site
elevation (meters); \code{strat} a stratification variable for sampling
with two levels (\code{"L"} and \code{"M"}) categorized by landscape
metrics at each site; \code{count}, the number of moose at each site;
\code{presence}, a factor that indicates whether at least one moose was
observed at each site (\code{0} implies no moose; \code{1} implies at
least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG:
3338) projected coordinate of each site. These data have
point-referenced support because each observation occurs at point
coordinates represented by a \code{POINT} geometry. Moose are most
prevalent in the southwestern and eastern parts of the Togiak region
(Figure\(~\)\ref{fig-moose-data}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> bin <- spglm(
+   formula = presence ~ elev + strat,
+   family = binomial,
+   data = moose,
+   spcov_type = "none"
+ )
R> 
R> bin_glm <- glm(
+   formula = presence ~ elev + strat,
+   family = binomial,
+   data = moose,
+ )
R> round(coef(bin), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
    -0.4247     -0.0003      0.8070 
\end{CodeOutput}
\begin{CodeInput}
R> round(coef(bin_glm), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
    -0.4247     -0.0003      0.8070 
\end{CodeOutput}
\begin{CodeInput}
R> round(sqrt(diag(vcov(bin))), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
     0.4208      0.0019      0.2906 
\end{CodeOutput}
\begin{CodeInput}
R> round(sqrt(diag(vcov(bin_glm))), digits = 4)
\end{CodeInput}
\begin{CodeOutput}
(Intercept)        elev      stratM 
     0.4208      0.0019      0.2906 
\end{CodeOutput}
\end{CodeChunk}

Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel}
helper functions are available and that each of the \code{spglm()}
models uses the same hierarchical likelihood
(Equation\textasciitilde{}\ref{eq-marginal}):

@@

The likelihood-based statistics AIC, AICc, BIC, and deviance are much
lower for the SPGLM, indicating a better fit relative to the GLM.

Instead of relying on likelihood-based statistics, models can be
compared using a cross-validation procedure
\citep{james2013introduction}. The \code{loocv()} function performs
leave-one-out cross validation, comparing the predicted mean (on the
response scale) to the observed response variable for each hold-out
observation, recomputing estimates of \(\boldsymbol{\beta}\) in each
iteration. Performing leave-one-out cross validation tends to be more
computationally efficient than fitting the model, as leave-one-out cross
validation requires only one set of products involving the inverse
covariance matrix (a primary computational burden), while fitting
traditional models requires these products for each optimization
iteration. After performing leave-one-out cross validation, statistics
like bias, mean-squared-prediction error (MSPE), and the square root of
MSPE (RMSPE) can be used to evaluate models:

@@

Both models have negligible bias, but the SPGLM has much lower MSPE and
RMSPE than the GLM, indicating the SPGLM predictions are far more
efficient. Both the liklelihood-based leave-one-out cross validation
metrics prefer the SPGLM to the GLM.

We may also want to compare the fit of two SPGLMs with different spatial
covariance structures. For example, we may want to compare the fit of an
SPGLM with the \texttt{"exponential"} spatial covariance to the fit of a
SPGLM with the \texttt{"gaussian"} spatial covariance:

@@

The SPGLM with the \code{"exponential"} spatial covariance
(\code{spbin}) has a slightly lower (better) deviance but slightly
higher (worse) AIC, AICc, and BIC than the SPGLM with the
\code{"gaussian"} spatial covariance (\code{spbin2}). Both SPGLMs have
similar leave-one-out cross validation metrics, though the SPGLM with
the \code{"gaussian"} spatial covariance has slightly lower (better)
RMSPE. For practical purposes, these models fit similarly.
